---
title: "Titanic"
author: "Winston Saunders"
date: "December 31, 2014"
output: html_document
---

###Introduction

This is a model for the kaggle Titanic dataset. The idea is to make a predictive model of survivor. Judging from leaderboards, 


```{r "get data", echo=9:10}

## this section gets the data and loads it into training and testing data files

        ##set working directory

        setwd("/Users/winstonsaunders/Documents/TitanicKaggle")

        ## get data sets
        training <- read.csv("train.csv", na.strings=c("NA", ""))
        testing <- read.csv("test.csv", na.strings=c("NA", ""))

        trainingRaw<-training
        testingRaw<-testing

```




##Model TWO

Base first model on Sex and Age per tutorial. Since there are many NA's in the data we replace those with the mean ages for male or females.  
  
The model is trained with a _Random Forest_. 

```{r "clean data1", echo=FALSE}

  ## Clean the data

        ## calculate the median age for the population, men and women
        meanAge<-mean(training$Age[is.na(training$Age)==FALSE])

        meanMaleAge<-mean(training$Age[training$Sex=="male"&is.na(training$Age)==FALSE])
        meanFemaleAge<-mean(training$Age[training$Sex=="female"&is.na(training$Age)==FALSE])

        ## substitute median ages based on sex
        
                training$Age[is.na(training$Age)&training$Sex=="male"]<-meanMaleAge
                training$Age[is.na(training$Age)&training$Sex=="female"]<-meanFemaleAge


                testing$Age[is.na(testing$Age)&testing$Sex=="male"]<-meanMaleAge
                testing$Age[is.na(testing$Age)&testing$Sex=="female"]<-meanFemaleAge


```
The mean male age is `r round(meanMaleAge, 1)` years and the mean female age is `r round(meanFemaleAge,1)` years. Here is a snapshot of the cleaned data.  Note the fractional year on substituted values. 

```{r}
training[1:6,c("Survived", "Age", "Sex")]
```

The first model looks at the following function

```{r "First model", warning=FALSE, echo=33, message=FALSE}

## This is a first simple model


      ## create data sub partition

        require(caret)
        set.seed(8675309)
        xx <- createDataPartition(y=training$Survived, p=0.80, list=FALSE)
   


        ##Define training and test sets

        train1<-training[xx,]
        test1<-training[-xx,]

        #train1 <- train1[,c("Survived","Sex", "Age")]
        #test1 <-   test1[,c("Survived","Sex", "Age")]


        ## simplify Age column by bucketing
        ## group in multiples of 5*n+1 (1, 6, 11, ...)
        ## train1$Age2 <- 16*round(train1$Age/16,0)
        ## test1$Age2 <- 16*round(test1$Age/16,0)

        ## set up training control

        control1 <- trainControl(method = "cv")

        ## model using random forest

        modelfunction<-Survived~Age+Sex

        model1 <- train(modelfunction, data = train1, method="rf", trControl = control1, prox=FALSE)

        prediction1 <- predict(model1, test1)
        



        ## assign to zero or one. 

        prediction1[prediction1<0.75]<-0
        prediction1[prediction1>=0.75]<-1
        prediction1<-as.factor(prediction1)


        t<-table(prediction1, test1$Survived)

        conf<-confusionMatrix(test1$Survived, prediction1)


```

Write final output

```{r "output", echo=11}
        
        ## Make model prediciton
        predictionT <- predict(model1, testing)

        ## factor into choices
        predictionT[predictionT<0.75]<-0
        predictionT[predictionT>=0.75]<-1
        Survived<-predictionT
        PassengerId<-testing$PassengerId

        ## Write output file
        Output<-cbind(PassengerId, Survived)
        write.csv(Output, "Model1.csv", row.names=FALSE)

```

```{r}
head(Output)
```

Here is the confusion table

```{r}
conf$table
```
 
The accuracy is `r round(100*conf$overall[1],1)`%.

